---
title: "Exploratory data analysis - insurance dataset"
output: html_notebook
---


##Set the working directory and reading the csv file
```{r}
setwd("C:/Users/Sowmya CR/Google Drive/jigsaw/EDA")

insurance <- read.csv("insurance.csv", stringsAsFactors = TRUE)
```


## Look at data dimensions and columns
```{r}
dim(insurance)
names(insurance)
# View(insurance)
```
About the dataset - 

##Summary statistics and data structure
```{r}
# Create data dictionary if not available. Identify numerical and categorical variables.
str(insurance)

# Summary statistics - identify outliers
summary(insurance)

```

```{r}
# Number of rows with outliers
library(dplyr)

insurance%>%filter(age>70)%>%nrow()


```


#Percentile breakup
```{r}


quantile(insurance$charges,p=c(76:100)/100)
```


## Univariate analysis of categorical variables - create tabulations & proportions to understand breakup amongst categories
```{r}
##Tabulations
table(insurance$sex)

table(insurance$smoker)

table(insurance$region)


#Proportions - prop.table


```


##Bivariate analysis of categorical variables - cross tabulations & proportions
```{r}
table(insurance$sex, insurance$region)
table(insurance$sex, insurance$smoker)
table(insurance$region, insurance$smoker)

##Proportions using prop.table


```

## Univariate analysis of numerical variables & identifying outliers
```{r}
box_age <- boxplot(insurance$age)
list <- box_age$out

box_bmi <- boxplot(insurance$bmi)
box_bmi$out
box_bmi$stats # five number summary

boxplot(insurance$charges)
```

##Treatment of outliers
```{r}
#gives the positions in the data where outliers are present
index<-which(insurance$age %in% list)

# Find values at the index position
insurance$age[index]

# Replace the outliers with third quartile
box_age$stats[5]

insurance$age[index] <- box_age$stats[5]

```


##Check for missing values
```{r}
colSums(is.na(insurance))

##replace missing values with mean
insurance$bmi[is.na(insurance$bmi)]<-mean(insurance$bmi,na.rm=TRUE)


#Replace missing values with median
insurance$bmi[is.na(insurance$bmi)]<- box_bmi[3]


```

##Aggregating continuous vs categorical
```{r}
aggregate(charges ~ smoker, data = insurance, FUN=mean)

aggregate(charges ~ sex, data = insurance, FUN=mean)

aggregate(charges ~ region, data = insurance, FUN=mean)

aggregate(charges ~ smoker+sex, data = insurance, FUN=mean)

agg_data <- aggregate(charges ~ smoker+sex+region, data = insurance, FUN=mean)

##Sort ascending & descending
agg_data[order(-agg_data$charges),]

```

##analysis with ggplot
```{r}
library(ggplot2)
ggplot(insurance,aes(x=charges)) + geom_histogram(aes(fill=as.factor(smoker)),position="dodge")
ggplot(insurance,aes(x=charges)) + geom_histogram(aes(fill=as.factor(smoker)),position="dodge") + facet_grid(region~sex)
```


```{r}
#box plots : 
ggplot(insurance,aes(y=charges,x=sex,fill=as.factor(sex)))+ geom_boxplot()
ggplot(insurance,aes(y=charges,x=sex,fill=as.factor(sex)))+ geom_boxplot() +facet_grid(region~smoker)
```

```{r}
#bmi vs charges
library(ggplot2)

insurance$bmi = as.numeric(insurance$bmi)
class(insurance$bmi)

p<-ggplot(insurance,aes(x=bmi,y=charges,color=smoker))
p+geom_point()+geom_smooth(aes(color=smoker))
p+geom_point()+facet_grid(.~sex)
p+geom_point()+facet_grid(smoker~region)
```

## Binning numeric variables
```{r}
##Creating age groups
range(insurance$age)
insurance$agegroup[insurance$age<25] <- "18-24"
insurance$agegroup[insurance$age >=25 & insurance$age < 35] <- "25-34"
insurance$agegroup[insurance$age >=35 & insurance$age < 45] <- "35-44"
insurance$agegroup[insurance$age >=45 & insurance$age < 55] <- "45-54"
insurance$agegroup[insurance$age >=55 & insurance$age < 65] <- "55-65"

class(insurance$agegroup)

insurance$agegroup = as.factor(insurance$agegroup)

library(plyr)

count(insurance$agegroup)
```


```{r}
#age vs charges
p<-ggplot(insurance,aes(x=charges,y=age,color=as.factor(sex)))
p+geom_point()
p+geom_point()+facet_grid(.~smoker)
p+geom_point()+facet_grid(smoker~region)
```

## Age group vs charges
```{r}
ggplot(data = insurance, aes(x=agegroup, y=charges))+ geom_boxplot(aes(fill=agegroup))+facet_grid(.~sex)

ggplot(data = insurance, aes(x=agegroup, y=charges))+ geom_boxplot(aes(fill=agegroup))+facet_grid(.~region)

ggplot(data = insurance, aes(x=agegroup, y=charges))+ geom_boxplot(aes(fill=agegroup))+facet_grid(.~smoker)

ggplot(data = insurance, aes(x=agegroup, y=charges))+ geom_boxplot(aes(fill=agegroup))+facet_grid(region~smoker)
```

##Children as categorical variable
```{r}
insurance$children <- as.factor(insurance$children)

count(insurance$children)

ggplot(data = insurance, aes(x=children, y=charges))+ geom_boxplot()
```


# exploring relationships among features: correlation matrix
```{r}
corrmat <- cor(insurance[c("age", "bmi", "children", "charges")])

library(corrplot)

#Visualise the correlation
corrplot(corrmat, method = "circle")
```

```{r}
# visualing relationships among features: scatterplot matrix
pairs(insurance[c("age", "bmi", "children", "charges")])
```

```{r}
#Data Preparation
#Adding new variables

# add a higher-order "age" term
insurance$age2 <- insurance$age^2

# add an indicator for BMI >= 30
insurance$bmi30 <- ifelse(insurance$bmi >= 30, 1, 0)
```


##Dummy variable creation
```{r}
library(dummies)



dummy_insurance=dummy.data.frame(insurance)

head(dummy_insurance)




```

##Normalising variables
```{r}
## scale variables
library(scales)

x<-scale(dummy_insurance)

head(x)

```


##Partitioning data into training and test sets
```{r}
## split into training and test sets
library(caTools)

set.seed(123)
split = sample.split(insurance$charges, SplitRatio = 0.7)
traindata = subset(insurance, split == TRUE)
testdata = subset(insurance, split == FALSE)
dim(traindata)

head(testdata)

```


